---
title: "Discovering Associations - Model 2"
output: pdf_document
author: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include = FALSE}
#load packages
library(vcd)
library(data.table)
library(tidyverse)
library(ggplot2)
library(nlme)
library(lme4)
library(bild)
library(geepack)
library(MuMIn)
```

## Binomial data of rose vase life days. 


Out of 1440 datapoints, we have 60 missing outcomes (total vase days), which is 4.1% of the data.
More data description, can copy some from simulation description.

```{r, include = FALSE}
#read in data
d <- fread('count_data_G6.csv') %>%
    mutate(garden=as.factor(garden),
         species=as.factor(species),
         rater=as.factor(rater),
         compound=as.factor(compound),
         subplotID=as.factor(subplotID),
         bushID=as.factor(bushID)) %>%
  na.omit()
```

Fit a longitudinal binary data predicting vase life. First need to transform the data into a binary outcome per day.

```{r lonitudinal_binary}

outmat<-matrix(nrow = nrow(d), ncol=max(d$tot.vase.days))

outmat[is.na(outmat)]<-1
for (i in 1:nrow(outmat)){
  outmat[i,c(d[i,tot.vase.days]:25)]<-0
  outmat[i,d[i,tot.vase.days]]<-1

}

outdf<-as.data.frame(outmat)
names(outdf)<-paste0("newVar_",names(outdf))

d_full<-d %>%
  bind_cols(outdf %>% as.data.frame()) %>%
  pivot_longer(contains("newVar"), names_to="day", values_to = "fresh") %>%
  mutate(day=as.numeric(gsub("newVar_V","",day)))

data_full_cc <- aggregate(fresh ~ compound + day, data = d_full, FUN = mean) %>%
  mutate(water=ifelse(compound==1,T,F))

ggplot(data = data_full_cc)+
  geom_hline(yintercept=0.5, linetype="dashed")+
  geom_line(aes(x = day, y = fresh, color = compound, size=water, alpha=water)) +
  scale_size_discrete(range=c(0.7,1.5),guide="none")+
  scale_alpha_discrete(range=c(0.65,1), guide="none")+
  theme_bw()+
  ylab("Proportion of fresh flowers")+
  xlab("Day")

#included this just to see if all datapoints are there..
#ggplot(data = data_full_cc)+geom_line(aes(x = day, y = fresh, color = compound))+facet_wrap(~compound)
```

We test Generalized Estimating Equations (GEEGLM function).

We Consider various specifications for the `working' correlation structure
Model1 =  Independence
Model2 =  Exchangeable
Model3 =  Auto-regressive

```{r}
gee_out1 <- geeglm(fresh ~ day + day:compound,
id=flowerID, data=d_full, family="binomial", corstr="independence")
summary(gee_out1)
```

For model2 the interaction effects do not show the compound effects observed in the graphs.

```{r}
gee_out2 <- geeglm(fresh ~ day + day:compound,
id=flowerID, data=d_full, family="binomial", corstr="exchangeable")
summary(gee_out2)
```

```{r}
gee_out3 <- geeglm(fresh ~ day + day:compound,
id=flowerID, data=d_full, family="binomial", corstr="ar1")
summary(gee_out3)
```

We compare the models using QIC: the auto-regressive working correlation seems to work better.

```{r}
QIC <- MuMIn::QIC
model.sel(gee_out1, gee_out2, gee_out3, rank = QIC)
```

We test more complex models. 

Model4: adding species.

```{r}
gee_out4 <- geeglm(fresh ~ species + day + day:compound,
id=flowerID, data=d_full, family="binomial", corstr="ar1")
```

Model5: adding garden.

```{r}
gee_out5 <- geeglm(fresh ~ garden + day + day:compound,
id=flowerID, data=d_full, family="binomial", corstr="ar1")
```

Model6: adding rater.

```{r}
gee_out6 <- geeglm(fresh ~ rater + day + day:compound,
id=flowerID, data=d_full, family="binomial", corstr="ar1")
summary(gee_out6)
```

Model7: adding rater and garden.

```{r}
gee_out7 <- geeglm(fresh ~ rater + garden + day + day:compound,
id=flowerID, data=d_full, family="binomial", corstr="ar1")
summary(gee_out7)
```

```{r}
model.sel(gee_out1, gee_out2, gee_out3, gee_out4, gee_out5, gee_out6, 
          gee_out7, rank = QIC)
```

We compare with a similar model corresponding to Model 7, but using an independence working correlation.

```{r}
gee_out8 <- geeglm(fresh ~ rater + day + day:compound,
id=flowerID, data=d_full, family="binomial", corstr="independence")
summary(gee_out8)
```

Adding compounds to the model provides a lower QIC but the effect is not anymore present in the interaction term.

```{r}
gee_out9 <- geeglm(fresh ~ compound + rater + garden + day + day:compound,
id=flowerID, data=d_full, family="binomial", corstr="ar1")
summary(gee_out9)
```


```{r}
model.sel(gee_out1, gee_out2, gee_out3, gee_out4, gee_out5, gee_out6, 
          gee_out7, gee_out8,gee_out9, rank = QIC)
```

## Results with model 7

```{r}
gee_coefficients<-as.data.frame(summary(gee_out7)$coefficients) %>%
  rownames_to_column("predictor") %>%
  filter(grepl("day:compound",predictor)) %>%
  dplyr::rename(pval=`Pr(>|W|)`) %>% 
  dplyr::mutate(one_sided_pval=ifelse(Estimate>0, pval/2, (1-pval/2)),
                p_adjusted=p.adjust(one_sided_pval, method="holm"),
         significant_higher=ifelse(p_adjusted<0.05, T, F))

ggplot(gee_coefficients %>%
         mutate(compound=factor(gsub("day:compound","",predictor), levels=2:15)), 
       aes(x=compound, y=Estimate, color=p_adjusted<0.05))+
  geom_hline(yintercept=0, linetype="dashed")+
  geom_errorbar(aes(ymin=Estimate - Std.err, ymax=Estimate +Std.err), width=0.2)+geom_point(size=3)+theme_bw()+
  scale_color_manual(values=c("grey","#6B0504"), name="Adjusted p<0.05")+coord_flip()+
  ylab("Betas compound:day")+
  xlab("Compound")+
  theme(legend.position = "top")
  #scale_()+
  #theme(axis.title.x = element_text(size=15))

```
