---
title: "Discovering Associations - Model 1"
output: pdf_document
author: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include = FALSE}
#load packages
library(vcd)
library(data.table)
library(tidyverse)
library(ggplot2)
library(nlme)
library(lme4)
library(bild)
```

## Count data of rose vase life days. 

### Poisson-mixed effects modelling. 

Out of 1440 datapoints, we have 60 missing outcomes (total vase days), which is 4.1% of the data.
More data description, can copy some from simulation description.

```{r, include = FALSE}
#read in data
d <- fread('count_data_G6.csv') %>%
    mutate(garden=as.factor(garden),
         species=as.factor(species),
         rater=as.factor(rater),
         compound=as.factor(compound),
         subplotID=as.factor(subplotID),
         bushID=as.factor(bushID)) %>%
  na.omit()
```

Checking the distribution of the Vase Days of the count data. It looks like there is a larger degree of overdispersion than the poisson model currently accounts for (nope, it's just conditional on the compound: the goodfit should be done per compound).


```{r, fig.width=6, fig.height=4 }

ggplot(d, aes(x=tot.vase.days, fill=compound))+geom_histogram(alpha=0.7, bins=25)+facet_wrap(~compound, ncol=5)+theme(legend.position = "none")

ggplot(d, aes(x=tot.vase.days, fill=compound))+geom_density(alpha=0.5)+theme(legend.position = "none")
#ggplot(d %>% filter(compound %in% c(1, 6, 14)), aes(x=tot.vase.days, fill=as.factor(compound)))+geom_density(alpha=0.3)

# d %>% group_by(compound) %>% 
#   dplyr::summarise(mean=mean(tot.vase.days),
#                    var=var(tot.vase.days)) %>%
#   ggplot(aes(x=mean, y=var))+
#   geom_abline(slope=1, intercept=0)+
#   geom_point()
ggplot(d, aes(x=compound,y=tot.vase.days, fill=compound))+geom_boxplot()+theme(legend.position = "none")


gf <- goodfit(d$tot.vase.days[d$compound==1], "poisson")
summary(gf)
plot(gf, type="standing", scale="raw") #I guess the fit would be better if we did it per compound
```

Initial Model (Model 0).

```{r, glmer}
#https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html#overdispersion
overdisp_fun <- function(model) {
rdf <- df.residual(model)
rp <- residuals(model,type="pearson")
Pearson.chisq <- sum(rp^2)
prat <- Pearson.chisq/rdf #pearson residuals over the residual degrees of freedom
pval <- pchisq(Pearson.chisq, df=rdf, lower.tail=FALSE)
c(chisq=Pearson.chisq,ratio=prat,rdf=rdf,p=pval)
}


#inspired by the first answer here for nested data formulation https://stats.stackexchange.com/questions/228800/crossed-vs-nested-random-effects-how-do-they-differ-and-how-are-they-specified
glmer_out <- glmer(tot.vase.days ~ compound + species + garden + (1|rater) + (1|subplotID/bushID), family=poisson(link = "log"), data=d)
summary(glmer_out)
-2*logLik(glmer_out)
AIC(glmer_out)
BIC(glmer_out)
overdisp_fun(glmer_out) #quite oki! no significance, yey. i think this might be more relevant than goodness of fit, because here we have all our predictors there that account for the variation

#https://fukamilab.github.io/BIO202/04-B-binary-data.html#glm_for_count_data
drop1(glmer_out, test = "Chi")



glmer_coefficients<-as.data.frame(summary(glmer_out)$coeff) %>%
  rownames_to_column("predictor") %>%
  filter(grepl("compound",predictor)) %>%
  dplyr::rename(pval=`Pr(>|z|)`) %>%
  #we want to have p-adjusted (Holm) values for one-sided test H.alt: lambda(compound)>lambda(water)
  dplyr::mutate(one_sided_pval=ifelse(`z value`>0, pval/2, (1-pval/2)),
         p_adjusted=p.adjust(one_sided_pval, method="holm"),
         significant_higher=ifelse(p_adjusted<0.05, T, F))

ggplot(glmer_coefficients %>%
         mutate(compound=factor(gsub("compound","",predictor), levels=2:15)), 
       aes(x=compound, y=Estimate, color=p_adjusted<0.05))+
  geom_hline(yintercept=0, linetype="dashed")+
  geom_errorbar(aes(ymin=Estimate - `Std. Error`, ymax=Estimate +`Std. Error`), width=0.2)+geom_point(size=3)+theme_bw()+
  scale_color_manual(values=c("grey","#85ADAC"))+coord_flip()+
  ylab(expression(beta["compound"]))
         
```

Conclusion: compounds 6 and 14 significantly increase rose vase days (mention estimates +- sd error, also backcalculated in days, alpha, one-sided Wald test, maybe the exact z and p-values, correction Holm).

### Models comparison. 

Model 1. Dropping Compounds.

```{r}
glmer_out1 <- glmer(tot.vase.days ~ species + garden + (1|rater) + (1|subplotID/bushID), family=poisson(link = "log"), data=d)
-2*logLik(glmer_out1)
AIC(glmer_out1)
BIC(glmer_out1)
LRT.1=2*(logLik(glmer_out)-logLik(glmer_out1))
c(LRT.1,1-pchisq(LRT.1,1))
```

Model 2. Dropping species.

```{r}
glmer_out2 <- glmer(tot.vase.days ~ compound + garden + (1|rater) + (1|subplotID/bushID), family=poisson(link = "log"), data=d)
-2*logLik(glmer_out2)
AIC(glmer_out2)
BIC(glmer_out2)
LRT.2=2*(logLik(glmer_out)-logLik(glmer_out2))
c(LRT.2,1-pchisq(LRT.2,1))
```

Model 3. Dropping garden.

```{r}
glmer_out3 <- glmer(tot.vase.days ~ compound + species + (1|rater) + (1|subplotID/bushID), family=poisson(link = "log"), data=d)
-2*logLik(glmer_out3)
AIC(glmer_out3)
BIC(glmer_out3)
LRT.3=2*(logLik(glmer_out)-logLik(glmer_out3))
c(LRT.3,1-pchisq(LRT.3,1))
```

Model 4.1 - Dropping bushID random effect.

```{r}
glmer_out4.1 <- glmer(tot.vase.days ~ compound + garden + species + (1|rater) + (1|subplotID), family=poisson(link = "log"), data=d)
-2*logLik(glmer_out4.1)
AIC(glmer_out4.1)
BIC(glmer_out4.1)
LRT.4.1=2*(logLik(glmer_out)-logLik(glmer_out4.1))
c(LRT.4.1,1-pchisq(LRT.4.1,1))
```

Model 4.2 - Dropping subplotID random effect

```{r}
glmer_out4.2 <- glmer(tot.vase.days ~ compound + garden + species + (1|rater) + (1|bushID), family=poisson(link = "log"), data=d)
-2*logLik(glmer_out4.2)
AIC(glmer_out4.2)
BIC(glmer_out4.2)
LRT.4.2=2*(logLik(glmer_out)-logLik(glmer_out4.2))
c(LRT.4.1,1-pchisq(LRT.4.1,1))
```

Dropping rater as random effect.

```{r}
glmer_out5 <- glmer(tot.vase.days ~ compound + garden + species + (1|subplotID/bushID), family=poisson(link = "log"), data=d)
-2*logLik(glmer_out5)
AIC(glmer_out5)
BIC(glmer_out5)
LRT.5=2*(logLik(glmer_out)-logLik(glmer_out5))
c(LRT.5,1-pchisq(LRT.5,1))
```


Model 6 is generated and compared to models 2 and 3.

```{r}
glmer_out6 <- glmer(tot.vase.days ~ compound + (1|subplotID/bushID) + (1|rater), family=poisson(link = "log"), data=d)
-2*logLik(glmer_out6)
AIC(glmer_out6)
BIC(glmer_out6)
LRT.6.2=2*(logLik(glmer_out2)-logLik(glmer_out6))
c(LRT.6.2,1-pchisq(LRT.6.2,1))
LRT.6.3=2*(logLik(glmer_out3)-logLik(glmer_out6))
c(LRT.6.3,1-pchisq(LRT.6.3,1))
```


Model 6 is compared with models lacking any of the random effects.
Model 7 without subplotID

```{r}
glmer_out7 <- glmer(tot.vase.days ~ compound + (1|bushID) + (1|rater), family=poisson(link = "log"), data=d)
-2*logLik(glmer_out7)
AIC(glmer_out7)
BIC(glmer_out7)
LRT.7=2*(logLik(glmer_out6)-logLik(glmer_out7))
c(LRT.7,1-pchisq(LRT.7,1))
```

Model 8 without bushID.

```{r}
glmer_out8 <- glmer(tot.vase.days ~ compound + (1|subplotID) + (1|rater), family=poisson(link = "log"), data=d)
-2*logLik(glmer_out8)
AIC(glmer_out8)
BIC(glmer_out8)
LRT.8=2*(logLik(glmer_out6)-logLik(glmer_out8))
c(LRT.8,1-pchisq(LRT.8,1))
```

Model 9 without rater.

```{r}
glmer_out9 <- glmer(tot.vase.days ~ compound + (1|subplotID/bushID), family=poisson(link = "log"), data=d)
-2*logLik(glmer_out9)
AIC(glmer_out9)
BIC(glmer_out9)
LRT.9=2*(logLik(glmer_out6)-logLik(glmer_out9))
c(LRT.9,1-pchisq(LRT.9,1))
```

### Results using model 6

```{r}
summary(glmer_out6)
glmer_coefficients<-as.data.frame(summary(glmer_out6)$coeff) %>%
  rownames_to_column("predictor") %>%
  filter(grepl("compound",predictor)) %>%
  dplyr::rename(pval=`Pr(>|z|)`) %>%
  #we want to have p-adjusted (Holm) values for one-sided test H.alt: lambda(compound)>lambda(water)
  dplyr::mutate(one_sided_pval=ifelse(`z value`>0, pval/2, (1-pval/2)),
         p_adjusted=p.adjust(one_sided_pval, method="holm"),
         significant_higher=ifelse(p_adjusted<0.05, T, F))

ggplot(glmer_coefficients %>%
         mutate(compound=factor(gsub("compound","",predictor), levels=2:15)), 
       aes(x=compound, y=Estimate, color=p_adjusted<0.05))+
  geom_hline(yintercept=0, linetype="dashed")+
  geom_errorbar(aes(ymin=Estimate - `Std. Error`, ymax=Estimate +`Std. Error`), width=0.2)+geom_point(size=3)+theme_bw()+
  scale_color_manual(values=c("grey","#85ADAC"))+coord_flip()+
  ylab(expression(beta["compound"]))
```