---
title: "Discovering Associations - Gaussian model selection"
output: pdf_document
author: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include = FALSE}
#load packages
library(vcd)
library(data.table)
library(tidyverse)
library(ggplot2)
library(nlme)
library(lme4)
library(bild)
library(lmerTest)
library(xtable)
library(multcomp)
```


## Gaussian data of flower width.

Gaussian outcome data. We received data from 180 flowers. This was distributed as 12 flowers for each compound for each of the 15 compounds. In each of those groups, there were 6 flowers per species and 6 grown in each garden. There were also 18 different subplots. The number of subplots is greater than the number of number of flowers per group. 

For each of the 18 flowers, we have measurements of the width of the flower over the course of 21 days. All measurements for all flowers were taken by a single rater.

Below I transform the data so that there is a row for each measurement of each flower on each day resulting in 3780 rows. 

```{r, message=FALSE, fig.width=8, fig.height=5}

g <- fread('gaussian_data_G6.csv') 
#summary(g) #there is only one rater, drop it
g<-g %>% dplyr::select(-Rater)


colnames(g)<-c("flowerID",0:20,"compound","type","garden","subplot")

dataG_long <- gather(g, days, width, "0":"20", factor_key=TRUE) %>%
      mutate(garden=as.factor(garden),
             type=as.factor(type),
             compound=as.factor(compound),
             subplot=as.factor(subplot),
             days=as.numeric(days),
             flowerID=as.factor(flowerID))

head(dataG_long)
```
I also added a column showing the change in the width of the flower so that we can see the change in width per day. It is worth noting that the width of the flower does not uniformly increase, instead it does fluctuate from day to day, decreasing occasionally. Also, there are quite a few missing measurements, we probably should have accounted for this in our sample size calculation?

```{r, include = FALSE}
dataG_combo <- data.table(flowerID = factor(), 
                          compound = factor(), 
                          type = factor(),
                          garden = factor(),
                          subplot = factor(),
                          days = integer(),
                          width = numeric(),
                          delta_width = numeric())

unique_flowerIDs <- unique(dataG_long$flowerID)
for (i in unique_flowerIDs) {
  dataG_subset <- dataG_long[dataG_long$flowerID == i,]
  dataG_subset$delta_width <- c(0,diff(dataG_subset$width))
  dataG_combo <- dplyr::union(dataG_combo, dataG_subset)
}
```


```{r}
p1=ggplot(dataG_combo[dataG_combo$days==20,], aes(x=compound,y=width, fill=compound))+geom_boxplot()+theme(legend.position = "none")+
  labs(y = "Width at day 20 (cm)", x= "Compound")
```



Below I plotted the individual flower width evolution (per coumpound) and the evolution of the mean flower width by compound. 

```{r}
p2=ggplot(data = dataG_combo)+
    geom_line(aes(x = days, y = width, color = flowerID)) +
  facet_wrap(~compound, ncol=5)+
  scale_size_discrete(range=c(0.8,1.5),guide="none")+
  scale_alpha_discrete(range=c(0.65,1), guide="none")+
  theme_bw()+theme(legend.position = "none")+
  ylab("Flower width (cm)")+
  xlab("Day")

cowplot::plot_grid(p1,p2, rel_widths = c(0.4,0.33), labels = c("A","B"), scale = 0.97)
ggsave("figures/mod3_overview.pdf", height=4, width=10)
```


```{r}
data_cc <- aggregate(width ~ compound + days, data = dataG_combo, FUN = mean) %>%
  mutate(water=ifelse(compound==1,T,F))

ggplot(data = data_cc)+
    geom_line(aes(x = days, y = width, color = compound, size=water, alpha=water)) +
  scale_size_discrete(range=c(0.8,1.5),guide="none")+
  scale_alpha_discrete(range=c(0.65,1), guide="none")+
  theme_bw()+
  ylab("Mean flower width")+
  xlab("Day")+labs(title = "Evolution of the mean flower width per coumpound")
```

The takeaway from this graph is that for each graph, the change in the Width of the flower is not the same for each of the Compounds. 

I fit a linear model to the gaussian outcome data where Compound, Type, Garden and Days are included as fixed effects, a compound and days interaction is included and subplot is included as a random effect. Rater is not included because we only have one rater. 

Initial model.

```{r}
lmer_out  <- lmer(width ~ compound + type + garden + days + compound:days +(1|subplot/flowerID), data=dataG_long)
summary(lmer_out)
```
```{r}
lmer_out.rs  <- lmer(width ~ compound + type + garden + days + compound:days +(1+days|subplot/flowerID), data=dataG_long)
summary(lmer_out.rs)
-2*logLik(lmer_out.rs)
AIC(lmer_out.rs)
BIC(lmer_out.rs)
anova(lmer_out,lmer_out.rs,refit=FALSE)
```

## Model selection

Dropping compound.

```{r}
lmer_out1 <- lmer(width ~ type + garden + days + compound:days + (1|subplot/flowerID), data=dataG_long)
summary(lmer_out1)
mod1<-as.data.frame(anova(lmer_out1,lmer_out, test="LRT",refit=FALSE)) %>%
  mutate(Model=c("Model 1","Full model"),
         dropped_predictor=c("Compound","-"))%>%
  relocate(c(Model, dropped_predictor))
```
Dropping species.

```{r}
lmer_out2 <- lmer(width ~ compound + garden + days + compound:days + (1|subplot/flowerID), data=dataG_long)
summary(lmer_out2)
mod2<-as.data.frame(anova(lmer_out2,lmer_out, test="LRT",refit=FALSE)) %>%
  mutate(Model=c("Model 2","Full model"),
         dropped_predictor=c("Species","-"))%>%
  relocate(c(Model, dropped_predictor))
```

Dropping garden.

```{r}
lmer_out3 <- lmer(width ~ compound + type + days + compound:days + (1|subplot/flowerID), data=dataG_long)
summary(lmer_out3)
mod3<-as.data.frame(anova(lmer_out3,lmer_out, test="LRT",refit=FALSE)) %>%
  mutate(Model=c("Model 3","Full model"),
         dropped_predictor=c("Garden","-"))%>%
  relocate(c(Model, dropped_predictor))
```

Dropping days.

```{r}
lmer_out4 <- lmer(width ~ compound + type + garden + compound:days + (1|subplot/flowerID), data=dataG_long)
mod4<-as.data.frame(anova(lmer_out4,lmer_out, test="LRT",refit=FALSE)) %>%
  mutate(Model=c("Model 4","Full model"),
         dropped_predictor=c("Days","-"))%>%
  relocate(c(Model, dropped_predictor))
```

Dropping compounds:days.

```{r}
lmer_out5 <- lmer(width ~ compound + type + garden + days + (1|subplot/flowerID), data=dataG_long)
summary(lmer_out5)
mod5<-as.data.frame(anova(lmer_out5,lmer_out, test="LRT",refit=FALSE)) %>%
  mutate(Model=c("Model 5","Full model"),
         dropped_predictor=c("Compound:Days","-"))%>%
  relocate(c(Model, dropped_predictor))
```


Dropping subplot random intercept.

```{r}
lmer_out6 <- lmer(width ~ compound + type + garden + days + compound:days + (1|flowerID), data=dataG_long)
summary(lmer_out6)
mod6<-as.data.frame(anova(lmer_out6,lmer_out, test="LRT",refit=FALSE)) %>%
  mutate(Model=c("Model 6","Full model"),
         dropped_predictor=c("Subplot","-"))%>%
  relocate(c(Model, dropped_predictor))
```

Dropping subplot flowerID intercept.

```{r}
lmer_out7 <- lmer(width ~ compound + type + garden + days + compound:days + (1|subplot), data=dataG_long)
summary(lmer_out7)
mod7<-as.data.frame(anova(lmer_out7,lmer_out, test="LRT",refit=FALSE)) %>%
  mutate(Model=c("Model 7","Full model"),
         dropped_predictor=c("FlowerID","-"))%>%
  relocate(c(Model, dropped_predictor))
```




```{r}
modselection<-bind_rows(mod1,mod2,mod3,mod4,mod5,mod6,mod7) %>%
  dplyr::rename("Dropped predictor"=dropped_predictor) %>%
  mutate(npar=as.factor(npar),
         Df=as.factor(Df))

print(xtable(modselection,digits=c(1,1,1,1,1,1,1,1,2,1,3)),
      include.rownames=FALSE, file="output_xtables/mod3_sel.tex")
```

```{r, lme}
p3<-plot(lmer_out1)
df <- data.frame(y = resid(lmer_out1))
p4 <- ggplot(df, aes(sample = y)) + stat_qq() + stat_qq_line()+theme_bw()+   xlab("Theoretical Quantiles") +
  ylab("Sample Quantiles")



cowplot::plot_grid(p3,p4, scale = c(1,0.9), labels = c("A","B"))
ggsave("figures/mod3_diag.pdf", height=4.8, width=8.3)

ctrl1<-summary(lmer_out1)$coeff['days',1]
lmer_coefficients<-as.data.frame(summary(lmer_out1)$coeff) %>%
  rownames_to_column("predictor") %>%
  filter(grepl("days",predictor)) %>%
  dplyr::rename(pval=`Pr(>|t|)`) %>% 
  dplyr::mutate(p_adjusted=p.adjust(pval, method="holm"),
         significant_higher=ifelse(p_adjusted<0.05, T, F),
         Estimate_width_increm=(Estimate+ctrl1)*20,
         Std.Error_width=0)
lmer_coefficients$p_adjusted[1]=1
lmer_coefficients$Estimate_width_increm[1]=lmer_coefficients$Estimate_width[1]-ctrl1*20
#Calculation of standard errors based on 
#https://stats.stackexchange.com/questions/446676/measuring-standard-error-of-two-or-more-coefficients-combined
lmer_coefficients$Std.Error_width[1]=lmer_coefficients$`Std. Error`[1]*20
a <- c(0,0,0,1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0)
for (val in 2:15){
  a2=a
  a2[val+3]=1
  lmer_coefficients$Std.Error_width[val]=sqrt(t(a2) %*% vcov(lmer_out1) %*% a2)*20
}

lmer_coefficients_plot=lmer_coefficients
lmer_coefficients_plot$predictor[1]="days:compound1"
p5<-ggplot(lmer_coefficients_plot %>%
         mutate(compound=factor(gsub("days:compound|","",predictor), levels=1:15)), 
       aes(x=compound, y=Estimate_width_increm, color=p_adjusted<0.05))+
  geom_hline(yintercept=ctrl1*20, linetype="dashed")+
  geom_errorbar(aes(ymin=Estimate_width_increm - Std.Error_width, ymax=Estimate_width_increm +Std.Error_width), width=0.2)+geom_point(size=1.8)+theme_bw()+
  scale_color_manual(values=c("grey","#6B0504"), name="Adjusted p<0.05")+coord_flip()+
  ylab("Rose width increment (cm) after 20 days post-harvest")+
  xlab("Compound")+
  theme(legend.position = "top")

ggsave("figures/mod3_outcome.pdf", plot=p5, height=4.8, width=8.3)
   
```

## Get Estimates Tables

Tables for the report are generated

```{r get_modtables}

lmer_mainTable<-as.data.frame(summary(lmer_out1)$coeff) %>%
  rownames_to_column("Predictor") 

lmer_mainTable=dplyr::select(lmer_mainTable, -df)

print(xtable(lmer_mainTable,digits=c(1,1,2,3,3,4)),
      include.rownames=FALSE, file="output_xtables/mod3_mainTable.tex")


#ranef(glmer_out7)
#https://stackoverflow.com/questions/8526681/extract-random-effect-variances-from-lme4-mer-model-object
lmer_randomTable<-as.data.frame(VarCorr(lmer_out1)) %>% dplyr::select(-var2)

print(xtable(lmer_randomTable,digits=c(1,1,1,3,3)),
      include.rownames=FALSE, file="output_xtables/mod3_randomTable.tex")

```

## Contrast

```{r}
wrap_glht_output<-function(glht_output){
  
  pq<-summary(glht_output)$test
  mtests <- cbind(pq$coefficients, pq$sigma, pq$tstat, pq$pvalues)
  error <- attr(pq$pvalues, "error")
  pname <- switch(glht_output$alternativ, 
                  less = paste("Pr(<", ifelse(glht_output$df ==0, "z", "t"), ")", sep = ""), 
                  greater = paste("Pr(>", ifelse(glht_output$df == 0, "z", "t"), ")", sep = ""), 
                  two.sided = paste("Pr(>|", ifelse(glht_output$df == 0, "z", "t"), "|)", sep = ""))     
  colnames(mtests) <- c("Estimate", "Std. Error", ifelse(glht_output$df ==0, "z value", "t value"), pname)
  return(mtests)
}

x<-glht(lmer_out1,  linfct=c("days:compound6 - days:compound14==0",
                             "days:compound6 - days:compound2==0",
                             "days:compound14 - days:compound2==0"))
mtests1<-wrap_glht_output(x)
print(xtable(mtests1,digits=c(1,2,3,3,4)), file="output_xtables/mod3_contrast_positive_compounds.tex")


x<-glht(lmer_out1,  linfct=c("days:compound4 - days:compound3=0",
                             "days:compound4 - days:compound8=0",
                             "days:compound3 - days:compound8=0"))
nmtests2<-wrap_glht_output(x)
print(xtable(nmtests2,digits=c(1,2,3,3,4)), file="output_xtables/mod3_contrast_negative_compounds.tex")
```