---
title: "Discovering Associations - Model 1"
output: pdf_document
author: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include = FALSE}
#load packages
library(vcd)
library(data.table)
library(tidyverse)
library(ggplot2)
library(nlme)
library(lme4)
#library(bild)
library(xtable)
library(multcomp)
library(emmeans)
```

## Count data of rose vase life days. 

### Poisson-mixed effects modelling. 

Out of 1440 datapoints, we have 60 missing outcomes (total vase days), which is 4.1% of the data.
More data description, can copy some from simulation description.

```{r, include = FALSE}
#read in data
d <- fread('count_data_G6.csv') %>%
    mutate(garden=as.factor(garden),
         species=as.factor(species),
         rater=as.factor(rater),
         compound=as.factor(compound),
         subplotID=as.factor(subplotID),
         bushID=as.factor(bushID)) %>%
  na.omit()
```

Checking the distribution of the Vase Days of the count data. It looks like there is a larger degree of overdispersion than the poisson model currently accounts for (nope, it's just conditional on the compound: the goodfit should be done per compound).


```{r, fig.width=6, fig.height=4 }
p1<-ggplot(d, aes(x=tot.vase.days, fill=compound))+geom_histogram(alpha=0.7, bins=25)+facet_wrap(~compound, ncol=5)+theme(legend.position = "none")+theme_bw()+theme(legend.position = "none", axis.title = element_text(size=11))+xlab("Total vase days")

p2<-ggplot(d, aes(x=compound,y=tot.vase.days, fill=compound))+geom_boxplot()+theme_bw()+
  theme(legend.position = "none")+ylab("Total vase days")+xlab("Compound")

cowplot::plot_grid(p1,p2, rel_widths = c(0.4,0.33), labels = c("A","B"), scale = 0.97)
ggsave("figures/mod1_overview.pdf", height=4, width=10)

#gf <- goodfit(d$tot.vase.days[d$compound==1], "poisson")
#summary(gf)
#plot(gf, type="standing", scale="raw") #I guess the fit would be better if we did it per compound
```

Initial Model (Model 0).

```{r, glmer}
#https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html#overdispersion
overdisp_fun <- function(model) {
rdf <- df.residual(model)
rp <- residuals(model,type="pearson")
Pearson.chisq <- sum(rp^2)
prat <- Pearson.chisq/rdf #pearson residuals over the residual degrees of freedom
pval <- pchisq(Pearson.chisq, df=rdf, lower.tail=FALSE)
c(chisq=Pearson.chisq,ratio=prat,rdf=rdf,p=pval)
}
#inspired by the first answer here for nested data formulation https://stats.stackexchange.com/questions/228800/crossed-vs-nested-random-effects-how-do-they-differ-and-how-are-they-specified
glmer_out <- glmer(tot.vase.days ~ compound + species + garden + (1|rater) + (1|subplotID/bushID), family=poisson(link = "log"), data=d)
summary(glmer_out)
-2*logLik(glmer_out)
AIC(glmer_out)
BIC(glmer_out)
overdisp_fun(glmer_out) #quite oki! no significance, yey. i think this might be more relevant than goodness of fit, because here we have all our predictors there that account for the variation
#https://fukamilab.github.io/BIO202/04-B-binary-data.html#glm_for_count_data
drop1(glmer_out, test = "Chi")
glmer_coefficients<-as.data.frame(summary(glmer_out)$coeff) %>%
  rownames_to_column("predictor") %>%
  filter(grepl("compound",predictor)) %>%
  dplyr::rename(pval=`Pr(>|z|)`) %>%
  #we want to have p-adjusted (Holm) values for one-sided test H.alt: lambda(compound)>lambda(water)
  dplyr::mutate(one_sided_pval=ifelse(`z value`>0, pval/2, (1-pval/2)),
         p_adjusted=p.adjust(one_sided_pval, method="holm"),
         significant_higher=ifelse(p_adjusted<0.05, T, F))
ggplot(glmer_coefficients %>%
         mutate(compound=factor(gsub("compound","",predictor), levels=2:15)), 
       aes(x=compound, y=Estimate, color=p_adjusted<0.05))+
  geom_hline(yintercept=0, linetype="dashed")+
  geom_errorbar(aes(ymin=Estimate - `Std. Error`, ymax=Estimate +`Std. Error`), width=0.2)+geom_point(size=3)+theme_bw()+
  scale_color_manual(values=c("grey","#85ADAC"))+coord_flip()+
  ylab(expression(beta["compound"]))
         
```

Conclusion: compounds 6 and 14 significantly increase rose vase days (mention estimates +- sd error, also backcalculated in days, alpha, one-sided Wald test, maybe the exact z and p-values, correction Holm).

### Models comparison. 

Model 1. Dropping Compounds.

```{r}
glmer_out1 <- glmer(tot.vase.days ~ species + garden + (1|rater) + (1|subplotID/bushID), family=poisson(link = "log"), data=d)
# -2*logLik(glmer_out1)
# AIC(glmer_out1)
# BIC(glmer_out1)
# LRT.1=2*(logLik(glmer_out)-logLik(glmer_out1))
# c(LRT.1,1-pchisq(LRT.1,1))

mod1<-as.data.frame(anova(glmer_out1,glmer_out, test="LRT")) %>%
  mutate(Model=c("Model 1","Full model"),
         dropped_predictor=c("Compound","-"))%>%
  relocate(c(Model, dropped_predictor))

```

Model 2. Dropping species.

```{r}
glmer_out2 <- glmer(tot.vase.days ~ compound + garden + (1|rater) + (1|subplotID/bushID), family=poisson(link = "log"), data=d)

mod2<-as.data.frame(anova(glmer_out2,glmer_out, test="LRT")) %>%
  mutate(Model=c("Model 2","Full model"),
         dropped_predictor=c("Species","-"))%>%
  relocate(c(Model, dropped_predictor))

```

Model 3. Dropping garden.

```{r}
glmer_out3 <- glmer(tot.vase.days ~ compound + species + (1|rater) + (1|subplotID/bushID), family=poisson(link = "log"), data=d)

mod3<-as.data.frame(anova(glmer_out3,glmer_out, test="LRT")) %>%
  mutate(Model=c("Model 3","Full model"),
         dropped_predictor=c("Garden","-"))%>%
  relocate(c(Model, dropped_predictor))
```

Model 4.1 - Dropping bushID random effect.

```{r}
glmer_out4 <- glmer(tot.vase.days ~ compound + garden + species + (1|rater) + (1|subplotID), family=poisson(link = "log"), data=d)

mod4<-as.data.frame(anova(glmer_out4,glmer_out, test="LRT")) %>%
  mutate(Model=c("Model 4","Full model"),
         dropped_predictor=c("BushID","-"))%>%
  relocate(c(Model, dropped_predictor))

```

Model 4.2 - Dropping subplotID random effect

```{r}
glmer_out5 <- glmer(tot.vase.days ~ compound + garden + species + (1|rater) + (1|bushID), family=poisson(link = "log"), data=d)

mod5<-as.data.frame(anova(glmer_out5,glmer_out, test="LRT")) %>%
  mutate(Model=c("Model 5","Full model"),
         dropped_predictor=c("subplotID","-"))%>%
  relocate(c(Model, dropped_predictor))

```

Dropping rater as random effect.

```{r}
glmer_out6 <- glmer(tot.vase.days ~ compound + garden + species + (1|subplotID/bushID), family=poisson(link = "log"), data=d)

mod6<-as.data.frame(anova(glmer_out6, glmer_out,test="LRT")) %>%
  mutate(Model=c("Model 6","Full model"),
         dropped_predictor=c("Rater","-"))%>%
  relocate(c(Model, dropped_predictor))

```


Model 7 (dropping Species and Garden) is generated and compared to models 2 (dropping Species) and 3 (dropping Garden).

```{r}
glmer_out7 <- glmer(tot.vase.days ~ compound + (1|subplotID/bushID) + (1|rater), family=poisson(link = "log"), data=d)
# -2*logLik(glmer_out6)
# AIC(glmer_out6)
# BIC(glmer_out6)
# LRT.6.2=2*(logLik(glmer_out2)-logLik(glmer_out6))
# c(LRT.6.2,1-pchisq(LRT.6.2,1))
# LRT.6.3=2*(logLik(glmer_out3)-logLik(glmer_out6))
# c(LRT.6.3,1-pchisq(LRT.6.3,1))


mod7.1<-as.data.frame(anova(glmer_out7,glmer_out2, test="LRT")) %>%
  mutate(Model=c("Model 7","Model 2"),
         dropped_predictor=c("Species, Garden","Species")) %>%
  relocate(c(Model, dropped_predictor))


mod7.2<-as.data.frame(anova(glmer_out7,glmer_out3, test="LRT")) %>%
  mutate(Model=c("Model 7","Model 3"),
         dropped_predictor=c("Species, Garden","Garden")) %>%
  relocate(c(Model, dropped_predictor))

```


Model 7 is compared with models lacking any of the random effects.
Model 8 without subplotID

```{r}
glmer_out8 <- glmer(tot.vase.days ~ compound + (1|bushID) + (1|rater), family=poisson(link = "log"), data=d)
# -2*logLik(glmer_out7)
# AIC(glmer_out7)
# BIC(glmer_out7)
# LRT.7=2*(logLik(glmer_out6)-logLik(glmer_out7))
# c(LRT.7,1-pchisq(LRT.7,1))

mod8<-as.data.frame(anova(glmer_out8,glmer_out7, test="LRT")) %>%
  mutate(Model=c("Model 8","Model 7"),
         dropped_predictor=c("Species, Garden, subplotID","Species, Garden")) %>%
  relocate(c(Model, dropped_predictor))

```

Model 9 without bushID.

```{r}
glmer_out9 <- glmer(tot.vase.days ~ compound + (1|subplotID) + (1|rater), family=poisson(link = "log"), data=d)
# -2*logLik(glmer_out8)
# AIC(glmer_out8)
# BIC(glmer_out8)
# LRT.8=2*(logLik(glmer_out6)-logLik(glmer_out8))
# c(LRT.8,1-pchisq(LRT.8,1))

mod9<-as.data.frame(anova(glmer_out9,glmer_out7, test="LRT")) %>%
  mutate(Model=c("Model 9","Model 7"),
         dropped_predictor=c("Species, Garden, bushID","Species, Garden")) %>%
  relocate(c(Model, dropped_predictor))

```

Model 10 without rater.

```{r}
glmer_out10 <- glmer(tot.vase.days ~ compound + (1|subplotID/bushID), family=poisson(link = "log"), data=d)
# -2*logLik(glmer_out9)
# AIC(glmer_out9)
# BIC(glmer_out9)
# LRT.9=2*(logLik(glmer_out6)-logLik(glmer_out9))
# c(LRT.9,1-pchisq(LRT.9,1))

mod10<-as.data.frame(anova(glmer_out10,glmer_out7, test="LRT")) %>%
  mutate(Model=c("Model 10","Model 7"),
         dropped_predictor=c("Species, Garden, Rater","Species, Garden")) %>%
  relocate(c(Model, dropped_predictor))
```

```{r modsel}

modselection<-bind_rows(mod1,mod2,mod3,mod4,mod5,mod6,mod7.1,mod7.2,mod8,mod9,mod10) %>%
  dplyr::rename("Dropped predictor"=dropped_predictor) %>%
  mutate(npar=as.factor(npar),
         Df=as.factor(Df))

print(xtable(modselection,digits=c(1,1,1,1,1,1,1,1,2,1,3)),
      include.rownames=FALSE, file="output_xtables/modelselection1.tex") #only added additional hlines after this

```


### Results using model 7

```{r}

p3<-plot(glmer_out7)
summary(glmer_out7)

#this is where emmeans package comes in
glmer1.rg <- ref_grid(glmer_out7)
glmer1.emm.s <- emmeans(glmer1.rg, "compound")

backcalc<-confint(glmer1.emm.s, level = .9, type = "response") %>%
  mutate(predictor=paste0("compound",compound))


# ctrl<-exp(summary(glmer_out)$coeff)[1,1]
# comp6<-exp(summary(glmer_out)$coeff)[6,1]
# comp14<-exp(summary(glmer_out)$coeff)[14,1]
# ctrl*comp6-ctrl
# ctrl*comp14-ctrl
ctrl1_noExp<-summary(glmer_out7)$coeff[1,1]
ctrl1<-exp(summary(glmer_out7)$coeff)[1,1]
ctrl1_stderr<-exp(summary(glmer_out7)$coeff)[1,2]


glmer_coefficients<-as.data.frame(summary(glmer_out7)$coeff) %>%
  rownames_to_column("predictor") %>%
  filter(grepl("compound",predictor)) %>%
  dplyr::rename(pval=`Pr(>|z|)`) %>%
  #we want to have p-adjusted (Holm) values for one-sided test H.alt: lambda(compound)>lambda(water)
  dplyr::mutate(one_sided_pval=ifelse(`z value`>0, pval/2, (1-pval/2)),
         p_adjusted=p.adjust(one_sided_pval, method="holm"),
         significant_higher=ifelse(p_adjusted<0.05, T, F),
         Estimate_days=exp(Estimate)*ctrl1,
         Std.Error_days=exp(`Std. Error`)*ctrl1_stderr) %>%
  left_join(backcalc)

#recheck backcalculation
cor(glmer_coefficients$Estimate_days,glmer_coefficients$rate)


pCIs<-ggplot(glmer_coefficients %>%
         mutate(compound=factor(gsub("compound","",predictor), levels=2:15)), 
       aes(x=compound, y=Estimate_days, color=p_adjusted<0.05))+
  geom_hline(yintercept=ctrl1, linetype="dashed")+
  geom_errorbar(aes(ymin=asymp.LCL, ymax=asymp.UCL), width=0.2)+geom_point(size=3)+theme_bw()+
  scale_color_manual(values=c("grey","#6B0504"), name="Adjusted p<0.05")+coord_flip()+
  ylab("Rose vase life in days")+
  xlab("Compound")+
  theme(legend.position = "top")


cowplot::plot_grid(
ggplot(glmer_coefficients %>%
         mutate(compound=factor(gsub("compound","",predictor), levels=2:15)), 
       aes(x=compound, y=Estimate_days, color=p_adjusted<0.05))+
  geom_hline(yintercept=ctrl1, linetype="dashed")+
  geom_errorbar(aes(ymin=Estimate_days - SE, ymax=Estimate_days + SE), width=0.2)+geom_point(size=3)+theme_bw()+
  scale_color_manual(values=c("grey","#6B0504"), name="Adjusted p<0.05")+coord_flip()+
  ylab("Rose vase life in days")+
  xlab("Compound")+
  theme(legend.position = "top"),


ggplot(glmer_coefficients %>%
         mutate(compound=factor(gsub("compound","",predictor), levels=2:15)), 
       aes(x=compound, y=Estimate_days, color=p_adjusted<0.05))+
  geom_hline(yintercept=ctrl1, linetype="dashed")+
  geom_errorbar(aes(ymin=Estimate_days - Std.Error_days, ymax=Estimate_days + Std.Error_days), width=0.2)+geom_point(size=3)+theme_bw()+
  scale_color_manual(values=c("grey","#6B0504"), name="Adjusted p<0.05")+coord_flip()+
  ylab("Rose vase life in days")+
  xlab("Compound")+
  theme(legend.position = "top")
)


p4<-ggplot(glmer_coefficients %>%
         mutate(compound=factor(gsub("compound","",predictor), levels=2:15)), 
       aes(x=compound, y=Estimate_days, color=p_adjusted<0.05))+
  geom_hline(yintercept=ctrl1, linetype="dashed")+
  geom_errorbar(aes(ymin=Estimate_days - SE, ymax=Estimate_days +SE), width=0.2)+geom_point(size=3)+theme_bw()+
  scale_color_manual(values=c("grey","#6B0504"), name="Adjusted p<0.05")+coord_flip()+
  ylab("Rose vase life in days")+
  xlab("Compound")+
  theme(legend.position = "top")
  #scale_()+
  #theme(axis.title.x = element_text(size=15))

cowplot::plot_grid(p3,p4, scale = c(0.93,0.99), labels = c("A","B"))
ggsave("figures/mod1_diagn_outcome.pdf", height=4.8, width=8.3)


ggsave("figures/mod1_diagnostics.pdf",cowplot::plot_grid(p3), height=3, width=4.5)

ggsave("figures/mod1_outcome.pdf",p4, height=5, width=4.5)


```

## Contrasts
Now do contrasts of the promising compounds.


```{r contr}
#https://stats.stackexchange.com/questions/60352/comparing-levels-of-factors-after-a-glm-in-r
#https://stat.ethz.ch/~meier/teaching/anova/contrasts-and-multiple-testing.html

wrap_glht_output<-function(glht_output){
  
  pq<-summary(glht_output)$test
  mtests <- cbind(pq$coefficients, pq$sigma, pq$tstat, pq$pvalues)
  error <- attr(pq$pvalues, "error")
  pname <- switch(glht_output$alternativ, 
                  less = paste("Pr(<", ifelse(glht_output$df ==0, "z", "t"), ")", sep = ""), 
                  greater = paste("Pr(>", ifelse(glht_output$df == 0, "z", "t"), ")", sep = ""), 
                  two.sided = paste("Pr(>|", ifelse(glht_output$df == 0, "z", "t"), "|)", sep = ""))     
  colnames(mtests) <- c("Estimate", "Std. Error", ifelse(glht_output$df ==0, "z value", "t value"), pname)
  return(mtests)
}

summary(glmer_out7)
x<-glht(glmer_out7,  linfct=mcp(compound=c("`6` - `14` == 0")))   

mtests1<-wrap_glht_output(x)

xtable(mtests1, digits=4)
print(xtable(mtests1,digits=c(1,2,3,3,4)), file="output_xtables/mod1_contrast_compound6vs14.tex")

```

```{r get_modtables}

rawCI<-confint(glmer1.emm.s, calc = c(n = ~.wgt.)) %>%
  mutate(Predictor=paste0("compound",compound),
         Predictor=ifelse(compound==1,"(Intercept)",Predictor))

glmer_mainTable<-as.data.frame(summary(glmer_out7)$coeff) %>%
  rownames_to_column("Predictor") %>%
  left_join(rawCI %>% dplyr::select(Predictor,n)) %>%
  mutate(lowerCI=Estimate-qt(0.9,df=n-1)*`Std. Error`,
         upperCI=Estimate+qt(0.9,df=n-1)*`Std. Error`, #use 0.9 because we are doing a one sided test here
         n=as.factor(n)) 
  
print(xtable(glmer_mainTable,digits=c(1,1,2,3,3,4,1,3,3)),
      include.rownames=FALSE, file="output_xtables/mod1_mainTable.tex")


#ranef(glmer_out7)
#https://stackoverflow.com/questions/8526681/extract-random-effect-variances-from-lme4-mer-model-object
glmer_randomTable<-as.data.frame(VarCorr(glmer_out7)) %>% dplyr::select(-var2)

print(xtable(glmer_randomTable,digits=c(1,1,1,3,3)),
      include.rownames=FALSE, file="output_xtables/mod1_randomTable.tex")

```



