---
title: "Discovering Associations - Model 2"
output: pdf_document
author: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include = FALSE}
#load packages
library(vcd)
library(data.table)
library(tidyverse)
library(tibble)
library(ggplot2)
library(nlme)
library(lme4)
library(bild)
library(geepack)
library(MuMIn)
library(xtable)
```

## Binomial data of rose vase life days. 


Out of 1440 datapoints, we have 60 missing outcomes (total vase days), which is 4.1% of the data.
More data description, can copy some from simulation description.

```{r, include = FALSE}
#read in data
d <- fread('count_data_G6.csv') %>%
    mutate(garden=as.factor(garden),
         species=as.factor(species),
         rater=as.factor(rater),
         compound=as.factor(compound),
         subplotID=as.factor(subplotID),
         bushID=as.factor(bushID)) %>%
  na.omit()
```

Fit a longitudinal binary data predicting vase life. First need to transform the data into a binary outcome per day.

```{r lonitudinal_binary}

outmat<-matrix(nrow = nrow(d), ncol=max(d$tot.vase.days))

outmat[is.na(outmat)]<-1
for (i in 1:nrow(outmat)){
  outmat[i,c(d[i,tot.vase.days]:25)]<-0
  outmat[i,d[i,tot.vase.days]]<-1

}

outdf<-as.data.frame(outmat)
names(outdf)<-paste0("newVar_",names(outdf))

d_full<-d %>%
  bind_cols(outdf %>% as.data.frame()) %>%
  pivot_longer(contains("newVar"), names_to="day", values_to = "fresh") %>%
  mutate(day=as.numeric(gsub("newVar_V","",day)))

data_full_cc <- aggregate(fresh ~ compound + day, data = d_full, FUN = mean) %>%
  mutate(water=ifelse(compound==1,T,F))

ggplot(data = data_full_cc)+
  geom_hline(yintercept=0.5, linetype="dashed")+
  geom_line(aes(x = day, y = fresh, color = compound, size=water, alpha=water)) +
  scale_size_discrete(range=c(0.7,1.5),guide="none")+
  scale_alpha_discrete(range=c(0.65,1), guide="none")+
  theme_bw()+
  ylab("Proportion of fresh flowers")+
  xlab("Day")
ggsave("figures/mod2_overview.pdf", height=4.5, width=6)

#included this just to see if all datapoints are there..
#ggplot(data = data_full_cc)+geom_line(aes(x = day, y = fresh, color = compound))+facet_wrap(~compound)
```

We test Generalized Estimating Equations (GEEGLM function).

We Consider various specifications for the `working' correlation structure
Model1 =  Independence
Model2 =  Exchangeable
Model3 =  Auto-regressive

```{r}
gee_out1 <- geeglm(fresh ~ day + day:compound,
id=flowerID, data=d_full, family="binomial", corstr="independence")
summary(gee_out1)
```

For model2 the interaction effects do not show the compound effects observed in the graphs.

```{r}
gee_out2 <- geeglm(fresh ~ day + day:compound,
id=flowerID, data=d_full, family="binomial", corstr="exchangeable")
summary(gee_out2)
```

```{r}
gee_out3 <- geeglm(fresh ~ day + day:compound,
id=flowerID, data=d_full, family="binomial", corstr="ar1")
summary(gee_out3)
```

We compare the models using QIC: the auto-regressive working correlation seems to work better.

```{r}
QIC <- MuMIn::QIC
model.sel(gee_out1, gee_out2, gee_out3, rank = QIC)
```

We test more complex models. 

Model4: adding species.

```{r}
gee_out4 <- geeglm(fresh ~ species + day + day:compound,
id=flowerID, data=d_full, family="binomial", corstr="ar1")
```

Model5: adding garden.

```{r}
gee_out5 <- geeglm(fresh ~ garden + day + day:compound,
id=flowerID, data=d_full, family="binomial", corstr="ar1")
```

Model6: adding rater.

```{r}
gee_out6 <- geeglm(fresh ~ rater + day + day:compound,
id=flowerID, data=d_full, family="binomial", corstr="ar1")
summary(gee_out6)
```

Model7: adding rater and garden.

```{r}
gee_out7 <- geeglm(fresh ~ rater + garden + day + day:compound,
id=flowerID, data=d_full, family="binomial", corstr="ar1")
summary(gee_out7)

```

```{r}
model.sel(gee_out1, gee_out2, gee_out3, gee_out4, gee_out5, gee_out6, 
          gee_out7, rank = QIC)
```

We compare with a similar model corresponding to Model 7, but using an independence working correlation.

```{r}
gee_out8 <- geeglm(fresh ~ rater +garden+ day + day:compound,
id=flowerID, data=d_full, family="binomial", corstr="independence")
summary(gee_out8)

```

Adding compounds to the model provides a lower QIC but the effect is not anymore present in the interaction term.

```{r}
gee_out9 <- geeglm(fresh ~ compound + rater + garden + day + day:compound,
id=flowerID, data=d_full, family="binomial", corstr="ar1")
summary(gee_out9)
```


```{r}
modelselection<-model.sel("model 1"=gee_out1, "model 2"=gee_out2, "model 3"=gee_out3,"model 4"= gee_out4, "model 5"=gee_out5, "model 6"=gee_out6,"model 7"=gee_out7,"model 8"=gee_out8,"model 9"=gee_out9, rank = QIC)
as.data.frame(modelselection)
dropped_predictor<-c("species","species,compound","species,compound","species,compound,garden",
                     "species,compound,rater","species,compound,garden,rater","compound,garden,rater",
                     "species,compound,garden,rater","species,compound,garden,rater")

modelselection[,c(3:7,9)]<-NULL
model2selection<-add_column(modelselection,"Model"= rownames(modelselection),"Dropped predictor"=dropped_predictor,.before = "(Intercept)")

print(xtable(model2selection,digits=c(1,1,1,1,1,1,1,1,1,1)),
      include.rownames=FALSE, file="output_xtables/model2selection.tex")

```

## Results with model 7

```{r}
gee_coefficients<-as.data.frame(summary(gee_out7)$coefficients) %>%
  rownames_to_column("predictor") %>%
  filter(grepl("day",predictor)) %>%
  dplyr::rename(pval=`Pr(>|W|)`) %>% 
  dplyr::mutate(one_sided_pval=ifelse(Estimate>0, pval/2, (1-pval/2)),
                p_adjusted=p.adjust(one_sided_pval, method="holm"),
         significant_higher=ifelse(p_adjusted<0.05, T, F))

gee_coefficients1<-gee_coefficients
gee_coefficients1[,c(5,8)]<-NULL
gee.out.tab<-gee_coefficients1%>%
  dplyr::rename("Pr(>z)" = "one_sided_pval","Holm ad-pvalue" = "p_adjusted","z value"="Wald","Std.Erorr"= "Std.err")

print(xtable(gee.out.tab,digits=c(1,1,3,3,3,3,3)),
      include.rownames=FALSE, file="output_xtables/gee_output.tex")


#ggplot(gee_coefficients %>%
#         mutate(compound=factor(gsub("day:compound","",predictor), levels=2:15)), 
#       aes(x=compound, y=Estimate, color=p_adjusted<0.05))+
#  geom_hline(yintercept=0, linetype="dashed")+
#  geom_errorbar(aes(ymin=Estimate - Std.err, ymax=Estimate +Std.err), #width=0.2)+geom_point(size=3)+theme_bw()+
#  scale_color_manual(values=c("grey","#6B0504"), name="Adjusted p<0.05")+coord_flip()+
#  ylab("Betas compound:day")+
#  xlab("Compound")+
#  theme(legend.position = "top")
#  #scale_()+
#  #theme(axis.title.x = element_text(size=15))


ctrl1<-summary(gee_out7)$coefficients['day',1]
gee2_coefficients<-as.data.frame(summary(gee_out7)$coefficients) %>%
  rownames_to_column("predictor") %>%
  filter(grepl("day",predictor)) %>%
  dplyr::rename(pval=`Pr(>|W|)`) %>% 
  dplyr::mutate(one_sided_pval=ifelse(Estimate>0, pval/2, (1-pval/2)),
                p_adjusted=p.adjust(one_sided_pval, method="holm"),
                significant_higher=ifelse(p_adjusted<0.05, T, F),
                Estimate_odds=exp(Estimate+ctrl1),
                Std.Error_odds=exp(Std.err+ctrl1)-exp(ctrl1),
                odds_ratio =Estimate_odds/exp(ctrl1))
gee2_coefficients$Estimate_odds[1]=exp(gee2_coefficients$Estimate[1])
gee2_coefficients$odds_ratio[1]= exp(gee2_coefficients$Estimate[1])/exp(ctrl1)

gee2_coefficients_plot=gee2_coefficients
gee2_coefficients_plot$predictor[1]="day:compound1"
p5<-ggplot(gee2_coefficients_plot %>%
         mutate(compound=factor(gsub("day:compound|","",predictor), levels=1:15)), 
       aes(x=compound, y=odds_ratio, color=p_adjusted<0.05))+
  geom_hline(yintercept=1, linetype="dashed")+
  geom_errorbar(aes(ymin=odds_ratio - Std.Error_odds, ymax=odds_ratio +Std.Error_odds), width=0.2)+geom_point(size=1.8)+theme_bw()+
  scale_color_manual(values=c("grey","#6B0504"), name="Adjusted p<0.05")+coord_flip()+
  ylab("Odds ratio off rose freshness compared to water")+
  xlab("Compound")+
  theme(legend.position = "top")
p5

```
