---
title: "Discovering Associations"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include = FALSE}
#load packages
library("lme4")
library(vcd)
library(data.table)
library(tidyr)
library(ggplot2)
library(nlme)
```

The data is loaded and manipulated

```{r}
#read in data
path <- 'count_data_G6.csv'
d <- fread(path)
g <- fread('gaussian_data_G6.csv')
```

```{r, include=FALSE}
gOld=g
```

```{r, include=FALSE, message=FALSE, fig.width=8, fig.height=5}
g$Flower_index<-factor(g$Flower_index)

colnames(g)<-c("Flower_index",0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,"Compound","Rater","Type","Garden","Subplot")

dataG_long <- gather(g, Days, Width, "0":"20", factor_key=TRUE)


dataG_long$Garden<-as.factor(dataG_long$Garden)
dataG_long$Type<-as.factor(dataG_long$Type)
dataG_long$Compound<-as.factor(dataG_long$Compound)
dataG_long$Subplot<-as.factor(dataG_long$Subplot)
dataG_long$Rater<-as.factor(dataG_long$Rater)
dataG_long$Days<-as.numeric(dataG_long$Days)

head(dataG_long)
```


```{r, include = FALSE}
dataG_combo <- data.table(Flower_index = character(), 
                          Compound = character(), 
                          Rater = character(),
                          Type = character(),
                          Garden = character(),
                          Subplot = character(),
                          Days = integer(),
                          Width = numeric(),
                          delta_Width = numeric())

unique_flower_ids <- unique(dataG_long$Flower_index)
for (i in unique_flower_ids) {
dataG_subset <- dataG_long[dataG_long$Flower_index == i,]
dataG_subset$delta_Width <- c(0,diff(dataG_subset$Width))
dataG_combo <- dplyr::union(dataG_combo, dataG_subset)
}
```


Below I plotted the mean width of the flower by day by compound on a given day. 

```{r}
data_cc <- aggregate(Width ~ Compound + Days, data = dataG_combo, FUN = mean)

plot <- ggplot(data = data_cc)+
    geom_line(aes(x = Days, y = Width, color = Compound)) 
   
plot
```

A similar plot is performed, but in this case as a boxplot of the width at the day 20. This allows a bit more clearly to see the number of the compound.


```{r}
gOld$Compound=as.factor(gOld$Compound)
ggplot(data=gOld, aes(x=Compound,y=T_20))+
  geom_boxplot()+ylab("Width at day 20 (cm)")
```

To compare the gaussian data with the Poisson data, we plot a similar boxplot for the total vase days.

```{r}
ggplot(data=d, aes(x=as.factor(compound),y=tot.vase.days))+
  geom_boxplot()+ylab("Total vase days")
```


A combined table is generated with the mean values of the two previous boxplots. 

```{r}
g.data.table=as.data.table(gOld)
g.data.table$Compound=as.factor(g.data.table$Compound)
g.data.table=na.omit(g.data.table, cols="T_20")
summary <- g.data.table[,.(MeanWidth=mean(T_20)),.(Compound)]
d.data.table=as.data.table(d)
d.data.table=na.omit(d.data.table, cols="tot.vase.days")
d.data.table$Compound=as.factor(d.data.table$compound)
summary2 <- d.data.table[,.(MedianVaseDays=median(tot.vase.days)),.(Compound)]
summary <- summary[summary2, on = .(Compound)]
summary
```

And then plotted

```{r}
ggplot(data = summary,aes(x = MeanWidth, y = MedianVaseDays,color = Compound))+
    geom_point()+xlab("Mean width at day 20 (cm)")+ylab("Median vase days")
```

A mixed model is fitted


```{r}
d$bushID<-as.factor(d$bushID)
d$compound<-as.factor(d$compound)
d$garden<-as.factor(d$garden)
d$species<-as.factor(d$species)
d$rater<-as.factor(d$rater)
m1 <-glmer(tot.vase.days~compound+species+garden+(1|rater)+(1|bushID)+(1|subplotID), family=poisson(link = "log"), data=d,na.action=na.omit)
summary(m1)
```


And we plot the compund effect based on the model and compare it with the width data.

```{r}
gOld2=as.data.table(gOld)
gOld2=na.omit(gOld2, cols="T_20")
gOld2=gOld2[,]
gSummary <- gOld2[,.(Mean=mean(T_20)),.(Compound)]
coefficients=summary(m1)$coefficients
coefficients=coefficients[c(1:15)]
coefficients=coefficients+coefficients[1]
coefficients[1]=coefficients[1]/2
gSummary$betas=coefficients
gSummary$Compound=as.factor(gSummary$Compound)
ggplot(data = gSummary,aes(x = Mean, y = betas,color = Compound))+
    geom_point()+xlab("Mean width at day 20 (cm)")+ylab("Compound effect based on the mixed model")
```
