---
title: "Analysis of the pilot data - PMHD - Group 6"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(hrbrthemes)
library(MASS)
```

## The data

The data is loaded and visualized with a histogram.

```{r}
pilot.data <- read.csv("G6.pilot.data.csv")
ggplot(data=pilot.data, aes(x=tot.vase.days)) +
  geom_histogram(color="#e9ecef", alpha=0.6, position = 'identity', binwidth = 1)
```

\newpage

## Plotting the histograms based on the different possible sources of variability

Garden

```{r}
ggplot(data=pilot.data, aes(x=tot.vase.days, fill=as.factor(garden))) +
  geom_histogram( color="#e9ecef", alpha=0.6, position = 'identity', binwidth = 1) +
  facet_wrap(~as.factor(garden),ncol=1)
```

\newpage

Species

```{r}
ggplot(data=pilot.data, aes(x=tot.vase.days, fill=as.factor(species))) +
  geom_histogram( color="#e9ecef", alpha=0.6, position = 'identity', binwidth = 1) +
  facet_wrap(~as.factor(species),ncol=1)
```

\newpage

Rater

```{r}
ggplot(data=pilot.data, aes(x=tot.vase.days, fill=as.factor(rater))) +
  geom_histogram( color="#e9ecef", alpha=0.6, position = 'identity', binwidth = 1) +
  facet_wrap(~as.factor(rater),ncol=1)
```

\newpage

## Estimation of the lambda parameter and confidence interval

The lambda parameter is estimated by fitting the pilot data to a poisson distribution.

```{r}
parms <- fitdistr(pilot.data$tot.vase.days, "poisson")
lambda.pilot <- parms$estimate
sd_x     <- parms$sd
confidence.interval.95 <- c(lambda.pilot + c(-1,1) * 1.96 * sd_x)
lambda.pilot
confidence.interval.95
```

\newpage

## Simulation to calculate the sample size

Simulation based on the code provided by the professor in the lecture of sample size calculation. 

```{r}
max.Sample.Size <- 300 # we evaluate all possible sample sizes up to 300
power <- rep(0,max.Sample.Size) # initialization of the variable
for (N in 1:max.Sample.Size){
    # setting the lambda parameters of the two populations
    lambda.control = lambda.pilot   # this value may come from pilot study
    lambda.treated = lambda.pilot+1 # value of the control group plus the effect size
    alpha = 0.05  # significance level
    
    # simulation of a high number of experiments and test
    numberSimulation <- 1000
    pval <- numeric(numberSimulation) # initialization of the variable 
    
    set.seed(1234) # set the seed for reproducible results
    for (i in 1:numberSimulation){
      # we simulate from Poisson distribution
      controlGroup <- rpois(N, lambda = lambda.control)
      treatedGroup <- rpois(N, lambda = lambda.treated)
      simData <- data.frame(response = c(controlGroup, treatedGroup), treatment = rep(c(0, 1), each = N))
      # we use a GLM model for Poisson regression to test effect of treatment
      pval[i] <- summary(glm(response ~ treatment, data = simData, family=poisson()))$coeff["treatment", "Pr(>|z|)"]
    }
    
    # Estimation of the power
    power[N]<-sum(pval < alpha)/numberSimulation
}
# plotting the results
sample.size <- 1:max.Sample.Size
plot(sample.size, power,xlab = "sample size")
abline(h=0.85, col="blue")
results <- data.frame(sample.size,power)
results
```


